---
title: "Gene_anno"
output: pdf_document
author: Tina Gorsek Sparovec
date: 26.1.22
---

```{r}

 if (!requireNamespace("BiocManager", quietly=TRUE))
     install.packages("BiocManager")
 BiocManager::install()
if (!requireNamespace("BiocManager", quietly=TRUE))

#BiocManager::install("topGO")
library(tidyverse)
 require(topGO)
require(org.Hs.eg.db)
library(janitor)
 library(topGO)
library(ALL)
library(pcaExplorer)
 BiocManager::install("hgu95av2.db")
data(ALL)
data(geneList)
affyLib <- paste(annotation(ALL), "db", sep = ".")
 library(package = affyLib, character.only = TRUE)
```
 
 
 
 ## TopGO vignette
 
 http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.515.8324&rep=rep1&type=pdf
 

```{r, eval=F}
sum(topDiffGenes(geneList))
head(topDiffGenes(geneList))
 sampleGOdata <- new("topGOdata",
                     description = "Simple session", ontology = "BP",
                     allGenes = geneList, geneSel = topDiffGenes,
                     nodeSize = 10,
                     annot = annFUN.db, affyLib = affyLib)

```
# TopGO on the RNAseq results
## Preparing the input vector

The input should be a named vector of DEG genes. Used is ENS_sig object created in
1.0_metadata_exploration. (changed on 11.7.22)

If using sig.res you need to change the code. 
```{r puting the genes as named vector}

genes<-ENS_sig #trying with ENS_sig instead of sig.genes

head(genes)

genes<-genes%>%
   #data.frame() %>% 
   #rownames_to_column(var="gene_names")%>%
   as_tibble()
genes<- genes %>%
  dplyr::select(c(gene_name, padj)) #selecting the columns gene and padj
  
all_genes_names <- genes$gene_name


# changing into the tibble with gene as rownames 
which(duplicated(genes$gene_name) == TRUE)
rownames(genes) <- make.names(genes$gene_name, unique=TRUE)
genes<-genes %>%  
  remove_rownames() 


#converting into dataframe 
genes<- as.data.frame(genes)


#get a factor with 2 levels--> named numeric vector

GeneList <- genes$padj #this is not correct 
GeneList<-as.numeric(GeneList) # converting doesnt work as.numeric
names (GeneList) <- all_genes_names #naming the vector 
head(GeneList)
head(geneList)
#https://www.biostars.org/p/2063/ 
#https://www.biostars.org/p/350710/ 


#https://www.biostars.org/p/471549/
selection <- function(allScore){ return(allScore < 0.05)} # function that returns TRUE/FALSE for p-values<0.05
allGO2genes_BP <- annFUN.org(whichOnto="BP", feasibleGenes=NULL, mapping="org.Hs.eg.db", ID="symbol")
GOdata_BP <- new("topGOdata",
  ontology="BP",  # you can change this to #NF #CP
  allGenes=GeneList,
  annot=annFUN.GO2genes,
  GO2genes=allGO2genes_BP,
  geneSel=selection,
  nodeSize=10) #creating the GOdata object which is then used for the GO analysis

allGO2genes_MF <- annFUN.org(whichOnto="MF", feasibleGenes=NULL, mapping="org.Hs.eg.db", ID="symbol")
GOdata_MF <- new("topGOdata",
  ontology="MF",  # you can change this to #NF #CP
  allGenes=GeneList,
  annot=annFUN.GO2genes,
  GO2genes=allGO2genes_MF,
  geneSel=selection,
  nodeSize=10)

allGO2genes_CC <- annFUN.org(whichOnto="CC", feasibleGenes=NULL, mapping="org.Hs.eg.db", ID="symbol")
GOdata_CC <- new("topGOdata",
  ontology="CC",  # you can change this to #NF #CP
  allGenes=GeneList,
  annot=annFUN.GO2genes,
  GO2genes=allGO2genes_CC,
  geneSel=selection,
  nodeSize=10)

#merging the GOdata
GOdata<-rbind(GOdata_BP,GOdata_MF, GOdata_CC)

#################### writing a loop --> doesnt work

All_genes<-lapply(Onto, annFUN.org, mapping="org.Hs.eg.db")
### creating a list of all genes for each ontology

#for loop--> not working
Onto<- c("BP", "MF", "CC")
GO_list<-list()
for (i in 1:length(Onto)){

  GO_res <- new("topGOdata",
  ontology= i,  # you can change this to #NF #CP
  allGenes=GeneList,
  annot=annFUN.GO2genes,
  GO2genes=All_genes[[i]],
  geneSel=selection,
  nodeSize=10)
  GO_list[[i]]<-GO_res
 }


# trying out with lapply
GO_res<- lapply(Onto, new, 
                allGenes=GeneList, 
                annot= annFUN.GO2genes, 
                GO2genes= All_genes, nodeSize=10, geneSel=selection)


GOdata <- new("topGOdata",
  ontology="MF",  # you can change this to #NF #CP
  allGenes=GeneList,
  annot=annFUN.GO2genes,
  GO2genes=All_genes[[2]],
  geneSel=selection,
  nodeSize=10)

GOdataMF <- new("topGOdata",
        ontology = "MF", # use biological process ontology
        allGenes = GeneList,
        geneSel=selection,
    annot = annFUN.GO2genes,
    GO2genes= All_genes[[2]])

```

##topGOtable
following the topgo vignette 
[topGO](https://bioconductor.org/packages/devel/bioc/vignettes/topGO/inst/doc/topGO.pdf)

### PCA explorer workflow

```{r}
#topGOtable with pcaExplorer ##NOT WORKING 
res<-ENS_sig

# tidying the df
ENS_sig$Row.names<-gsub("\\..*","",ENS_sig$Row.names)
rownames(ENS_sig)<-make.names(ENS_sig$Row.names, unique=T)

ENS_res$symbol <- mapIds(org.Hs.eg.db,
                            keys=row.names(ENS_res),
                            column="SYMBOL",
                            keytype="ENSEMBL",
                            multiVals="first") # P- getting symbols in 
ENS_res$entrez <- mapIds(org.Hs.eg.db,
                            keys=row.names(ENS_res),
                            column="ENTREZID",
                            keytype="ENSEMBL",
                            multiVals="first")
resOrdered <- as.data.frame(ENS_res[order(res$padj),])
head(resOrdered)

# extract DE genes
de_df <- resOrdered[resOrdered$padj < .05 & !is.na(resOrdered$padj),]
de_symbols <- de_df$symbol
#getting BG genes
bg_ids <- rownames(dds)[rowSums(counts(dds)) > 0]
bg_ids<-gsub("\\..*","",bg_ids)
bg_symbols <- mapIds(org.Hs.eg.db,
                     keys=bg_ids,
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")
GOdata_2<-topGOtable(de_symbols, bg_symbols,
                            ontology="BP", 
                            mapping = "org.Hs.eg.db",
                            geneID = "symbol") 
  

```
 
##Performing an enrichment with Fisher's exact test 
 You can actually use 2 different (F. exact test or Kolmogorov- Smirtnov test with computes enrichmet
 based on the gene count)
 
```{r}
#performing an enrichment 

resultFisher <- runTest(GOdata, algorithm = "classic", statistic = "fisher")
resultFisher
```
 testing the enrichment using KS test, with classic and elim method
 
```{r}
resultKS <- runTest(GOdata, algorithm = "classic", statistic = "ks")
resultKS.elim <- runTest(GOdata, algorithm = "elim", statistic = "ks", na.rm=T)
##### error not enough y data
## could be the problem with NA in ENS_sig

#### Solution in other vignette (link below), to get the elim results
test.stat <- new("elimCount", testStatistic = GOFisherTest,
                 name = "Fisher test", cutOff = 0.01)
resultElim <- getSigGroups(GOdata, test.stat)
#http://compdiag.molgen.mpg.de/ngfn/docs/2007/sep/topGO_Exercises.pdf 
```

## Analysis of results

analysis of most sig. GO terms and coresponding p-val
GenTable() returns a df containing topnodes GO term, for assessing the GO's term pval- score()
http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.515.8324&rep=rep1&type=pdf

```{r}

allRes <- GenTable(GOdata, classicFisher = resultFisher,
                    classicKS = resultKS, elimKS = resultElim, #with elim method from the second elim calculation (see solution)
                    orderBy = "elimKS", ranksOf = "classicFisher", topNodes = 50) # go for all of them and select what you want 
allRes #depends on the threshold that you use

 pValue.classic <- score(resultKS)
 pValue.elim <- score(resultElim)[names(pValue.classic)]
 gstat <- termStat(GOdata, names(pValue.classic))
 gSize <- gstat$Annotated / max(gstat$Annotated) * 4
 colMap <- function(x) {
  .col <- rep(rev(heat.colors(length(unique(x)))), time = table(x))
  return(.col[match(1:length(x), order(x))])
}
 gCol <- colMap(gstat$Significant)
 GOplot<-plot(pValue.classic, pValue.elim, xlab = "p-value classic", ylab = "p-value elim",
      pch = 19, cex = gSize, col = gCol)
 write.csv(allRes, "GOterms.csv")
```
references for comparison: https://www.frontiersin.org/articles/10.3389/fgene.2020.551495/full 

```{r visualisation of GO}
#--------Source: https://www.biostars.org/p/471549/
require(ggplot2)
library(scales)

goEnrichment$classicFisher<-as.numeric(goEnrichment$Fisher)
ntop <- 30
ggdata <- goEnrichment[1:ntop,]
ggdata$Term <- factor(ggdata$Term, levels = rev(ggdata$Term)) # fixes order
gg1 <- ggplot(ggdata,
  aes(x = Term, y = -log10(Fisher), size = -log10(Fisher), fill = -log10(Fisher))) +

  expand_limits(y = 1) +
  geom_point(shape = 21) +
  scale_size(range = c(2.5,12.5)) +
  scale_fill_continuous(low = 'royalblue', high = 'red4') +

  xlab('') + ylab('Enrichment score') +
  labs(
    title = 'GO Biological processes',
    subtitle = 'Top 30 terms ordered by Kolmogorov-Smirnov p-value',
    caption = 'Cut-off lines drawn at equivalents of p=0.05, p=0.01, p=0.001') +

  geom_hline(yintercept = c(-log10(0.05), -log10(0.01), -log10(0.001)),
    linetype = c("dotted", "longdash", "solid"),
    colour = c("black", "black", "black"),
    size = c(0.5, 1.5, 3)) +

  theme_bw(base_size = 24) +
  theme(
    legend.position = 'right',
    legend.background = element_rect(),
    plot.title = element_text(angle = 0, size = 16, face = 'bold', vjust = 1),
    plot.subtitle = element_text(angle = 0, size = 14, face = 'bold', vjust = 1),
    plot.caption = element_text(angle = 0, size = 12, face = 'bold', vjust = 1),

    axis.text.x = element_text(angle = 0, size = 12, face = 'bold', hjust = 1.10),
    axis.text.y = element_text(angle = 0, size = 12, face = 'bold', vjust = 0.5),
    axis.title = element_text(size = 12, face = 'bold'),
    axis.title.x = element_text(size = 12, face = 'bold'),
    axis.title.y = element_text(size = 12, face = 'bold'),
    axis.line = element_line(colour = 'black'),

    #Legend
    legend.key = element_blank(), # removes the border
    legend.key.size = unit(1, "cm"), # Sets overall area/size of the legend
    legend.text = element_text(size = 14, face = "bold"), # Text size
    title = element_text(size = 14, face = "bold")) +

  coord_flip()
```


```{r}
sel.go <- names(pValue.classic)[pValue.elim < pValue.classic]
 cbind(termStat(GOdata, sel.go),
       elim = pValue.elim[sel.go],
       classic = pValue.classic[sel.go])
 pdf("plotGO.pdf")
 showSigOfNodes(GOdata, score(resultKS.elim), firstSigNodes = 5, useInfo = 'all')
 dev.off()
```

# Annotation 
nr. 5) in the vignete
```{r}
BPterms <- ls(GOBPTerm)
head(BPterms)

```

```{r working with TopGodData}
 a <- genes(GOdata) ## obtain the list of genes
 head(a)
 numGenes(GOdata)
 
 # retrieving a score of specified set of genes
selGenes <- sample(a, 10)
 gs <- geneScore(GOdata, whichGenes = selGenes)
 print(gs)
 
 # unnamed vector or the scrore of all genes
 gs<-geneScore(GOdata, whichGenes = selGenes, use.names=F)
 print(gs)
 gs <- geneScore(GOdata, use.names = FALSE)
 str(gs)

#list of sig. genes
 sg <- sigGenes(GOdata)
str(sg)
numSigGenes(GOdata)

# updating the genes 
.geneList <- geneScore(GOdata, use.names = TRUE)
 GOdata ## more available genes
 GOdata <- updateGenes(GOdata, .geneList, topDiffGenes)
 GOdata ## the available genes are now the feasible genes
 graph(GOdata)                                              # ???is this number of nods7edges enough????
```


```{r}
ug <- usedGO(GOdata)
 head(ug)
 
 sel.terms <- sample(usedGO(GOdata), 10) #selecting 10 random GO terms to get the genes 
num.ann.genes <- countGenesInTerm(GOdata, sel.terms) ## the number of annotated genes > num.ann.genes
 ann.genes <- genesInTerm(GOdata, sel.terms) ## get the annotations
 head(ann.genes)
 
 #sel.term missing== you get all the GO terms
 ann.score <- scoresInTerm(GOdata, sel.terms)
 head(ann.score)
 ann.score <- scoresInTerm(GOdata, sel.terms, use.names = TRUE)
 head(ann.score)
 termStat(GOdata, sel.terms)
```
´


```{r}
#reactomePA
library(ReactomePA)
R_PA<-enrichPathway(
names(GeneList),
organism = "human"
) #returns errors 
```


To do:

We cont with likelihood ratio test, ~ for the model use also other factors that influence- you need even number of comparison (6=6)
variance partition package https://rdrr.io/bioc/variancePartition/f/inst/doc/variancePartition.pdf 

doing heatmap using the genes (done)

age clust by similarity

- topGOtable
- pathways and functions - make a table 


#topgotable
```{r}
#using res, anno, PAS 
dds_GO <- DESeq(dds)
res<- results(dds_GO)
ENS_res<- merge(as.data.frame(res), PASbackup, by.y="gene_name", by.x= "row.names")
ENS_res$gene_id<-gsub("\\..*","",ENS_res$gene_id)
genenames_GO<-mapIds(org.Hs.eg.db,
                     keys=ENS_res$gene_id, 
                     column = "SYMBOL",
                     keytype = "ENSEMBL")
annotation_GO <- data.frame(gene_name = genenames_GO, #some duplication 
                                row.names = rownames(ENS_res),
                                stringsAsFactors = FALSE)
head(annotation_GO)  

anno_df_orgdb <- get_annotation_orgdb(dds = dds_GO,
                                      orgdb_species = "org.Hs.eg.db",
                                      idtype = "ENSEMBL")

anno_df_biomart <- get_annotation(dds = dds,
                                  biomart_dataset = "hsapiens_gene_ensembl",
                                  idtype = "ensembl_gene_id")


res$symbol<- mapIds(org.Hs.eg.db,
                            keys= row.names(res),
                            column="SYMBOL",
                            keytype="ENSEMBL",
                            multiVals="first")
res$entrez <- mapIds(org.Hs.eg.db, ##problem 
                            keys=row.names(res),
                            column="ENTREZID",
                            keytype="ENSEMBL",
                            multiVals="first")
resOrdered <- as.data.frame(res[order(res$padj),])
head(resOrdered)

de_df <- resOrdered[resOrdered$padj < .05 & !is.na(resOrdered$padj),]
de_symbols <- de_df$symbol
# extract background genes
bg_ids <- rownames(dds_GO)[rowSums(counts(dds_GO)) > 0]
bg_symbols <- mapIds(org.Hs.eg.db,
                     keys=bg_ids,
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")
# run the function
topgoDE_GO <- topGOtable(de_symbols, bg_symbols,
                             ontology = "BP",
                             mapping = "org.Hs.eg.db",
                             geneID = "symbol")
#error NA in rownames
```

 