---
title: "Var_part.rmd"
output: pdf_document
author: Tina Gorsek Sparovec
date: february 2022
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Variance partitioning 

framework for understanding drivers of variation
in gene expression in experiments with complex designs. A
typical application would consider a dataset of gene expression from individuals
sampled in multiple tissues or multiple time points where the goal is to understand variation within versus between individuals and tissues. variancePartition
use a linear mixed model to partition the variance attributable to multiple variables in the data. 


Needs normalised expression values

3 components:
1) gene expression matrix
2) metadata
3) model
```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("variancePartition")
library("variancePartition")
#data used sig.res and anno metadata
```

```{r}
#example
data(varPartData)
form<-~Age+ (1|Individual) + (1| Tissue) + (1| Batch)
varPart<- fitExtractVarPartModel( geneExpr, form, info)
vp<-sortCols(varPart)
plotPercentBars(vp[1:10,])
plotVarPart(vp)
```

```{r}
dds<-estimateSizeFactors(dds) # tries to get rid of the variations between samples
expr<- rowSums(fpm(dds)>1) >= 0.5* ncol(dds)

#transformation, plus filter for the genes that pass the 0.05
quantlog<-log2(fpm(dds)[expr,] +1)

form <- ~(weight+ Grav + Para + sectio + Csec+ place) #which one is categorical, fixed or continuous
#uses rld table results for input
varPart_pas<- fitExtractVarPartModel(quantlog,form, anno) #problem
plotVarPart(varPart_pas)
?fitExtractVarPartModel
```
```{r}
# plot of varienace fractions for a subset of genes
plotPercentBars(varPart_pas[1:10,], col = c(ggColorHue(ncol(varPart_pas) - 1), "grey85"))

```
How to get the genes not the ensemble ID? 

```{r}
sortCols(varPart_pas)
plotVarPart( sortCols( varPart_pas ) )
```


