---
title: "Gene_anno"
author: "Tina Gorsek Sparovec"

date: 26.1.22
params:
 data: "../data"
 results: "../res"
 lib: "../lib"
 maxcores: 6
output:
 html_notebook:
 theme: spacelab
 highlight: textmate
 toc: yes
 number_sections: true
 toc_depth: 3
 toc_float: true
---

```{r}
library(tidyverse)
require(org.Hs.eg.db)
library(janitor)
library(topGO)
library(ALL)
library(pcaExplorer)
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("hgu95av2.db")

# Required data: from notebook 1.0_metadata_exploration

# (1) subtraction/subset of the results_PAS_Ctrl and results_CtrlMus_Ctrl 

ENS_sig
 
# (2) Background genes: genes of results_CtrlMus_Ctrl

ENS_bg
```
 
 
 
 ## TopGO vignette
 
 http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.515.8324&rep=rep1&type=pdf
 

```{r, eval=F}
data(ALL)
data(geneList)
affyLib <- paste(annotation(ALL), "db", sep = ".")
 library(package = affyLib, character.only = TRUE)
sum(topDiffGenes(geneList))
head(topDiffGenes(geneList))
 sampleGOdata <- new("topGOdata",
                     description = "Simple session", ontology = "BP",
                     allGenes = geneList, geneSel = topDiffGenes,
                     nodeSize = 10,
                     annot = annFUN.db, affyLib = affyLib)

```

# TopGO on the RNAseq results


The input should be a named vector of DEG genes. Used is ENS_sig object created in
1.0_metadata_exploration. (changed on 11.7.22)


```{r puting the genes as named vector}

genes <- ENS_sig |>
  rownames_to_column(var = "gene_id") |>
  pull(gene_id)

head(genes)

up_genes <- ENS_sig |>
  subset(padj < 0.05 & ENS_sig$log2FoldChange > 0) |>
  pull(Row.names)

dn_genes <- ENS_sig |>
  subset(padj < 0.05 & ENS_sig$log2FoldChange < 0) |>
  pull(Row.names)

# Extracting the background genes
bg_genes <- ENS_bg |>
  pull(Row.names)

```

##topGOtable

following the topgo vignette 
[topGO](https://bioconductor.org/packages/devel/bioc/vignettes/topGO/inst/doc/topGO.pdf)

```{r}
Res <- lapply(c("BP", "MF"), function(ontology) {
  topgode <- topGOtable(DEgenes = genes,
                        BGgenes = bg_genes,
                        ontology = ontology,
                        mapping = "org.Hs.eg.db",
                        geneID= "ENSEMBL"
                        )
  })


Res
```

 
##Performing an enrichment with Fisher's exact test 
 You can actually use 2 different (F. exact test or Kolmogorov- Smirtnov test with computes enrichmet
 based on the gene count)
 
```{r}
#performing an enrichment 

resultFisher <- runTest(GOdata_MF, algorithm = "classic", statistic = "fisher")
resultFisher
```
 testing the enrichment using KS test, with classic and elim method
 
```{r}
resultKS <- runTest(GOdata_MF, algorithm = "classic", statistic = "ks")
resultKS.elim <- runTest(GOdata_MF, algorithm = "elim", statistic = "ks", na.rm=T)
##### error not enough y data
## could be the problem with NA in ENS_sig

#### Solution in other vignette (link below), to get the elim results
test.stat <- new("elimCount", testStatistic = GOFisherTest,
                 name = "Fisher test", cutOff = 0.01)
resultElim <- getSigGroups(GOdata_MF, test.stat)
#http://compdiag.molgen.mpg.de/ngfn/docs/2007/sep/topGO_Exercises.pdf 
```

## Analysis of results

analysis of most sig. GO terms and coresponding p-val
GenTable() returns a df containing topnodes GO term, for assessing the GO's term pval- score()
http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.515.8324&rep=rep1&type=pdf

The elim method was design to be more conservative then the
classic method and therefore one expects the p-values returned by the former method 
are lower bounded by the p-values returned by the later method

```{r}

allRes <- GenTable(GOdata_MF, 
                  classicFisher = resultFisher,
                  classicKS = resultKS, 
                  elimKS = resultElim, #with elim method from the second elim calculation (see solution)
                  orderBy = "elimKS", ranksOf = "classicFisher", topNodes = 50) # go for all of them and select what you want 
allRes #depends on the threshold that you use
#### Result is a topNodes GO terms identified by the elim algorithm


## Accessing p-values from the GO object
 pValue.classic <- score(resultKS)
 pValue.elim <- score(resultElim)[names(pValue.classic)]
 gstat <- termStat(GOdata_MF, names(pValue.classic))
 gSize <- gstat$Annotated / max(gstat$Annotated) * 4
 colMap <- function(x) {
  .col <- rep(rev(heat.colors(length(unique(x)))), time = table(x))
  return(.col[match(1:length(x), order(x))])
 }
 
 gCol <- colMap(gstat$Significant)
plot(pValue.classic, pValue.elim, xlab = "p-value classic", ylab = "p-value elim",
      pch = 19, cex = gSize, col = gCol)
 
 
 write.csv(allRes, "GOterms.csv")
 
```

some pvalues will be less significant in elim compared to the classical. To idenify these we run the following command: 

```{r}
 
sel.go <- names(pValue.classic)[pValue.elim < pValue.classic]
 cbind(termStat(GOdata_MF, sel.go),
elim = pValue.elim[sel.go],
classic = pValue.classic[sel.go])
```
references for comparison with PAS: https://www.frontiersin.org/articles/10.3389/fgene.2020.551495/full 




```{r visualisation of GO}
#--------Source: https://www.biostars.org/p/471549/
require(ggplot2)
library(scales)

allRes$classicFisher <- as.numeric(allRes$classicFisher)
ntop <- 30
ggdata <- allRes[1:ntop,]
ggdata$Term <- factor(ggdata$Term, levels = rev(unique(ggdata$Term))) # fixes order, set as unique due to error
gg1 <- ggplot(ggdata,
  aes(x = Term, y = -log10(classicFisher), size = -log10(classicFisher), fill = -log10(classicFisher))) +

  expand_limits(y = 1) +
  geom_point(shape = 21) +
  scale_size(range = c(2.5,12.5)) +
  scale_fill_continuous(low = 'royalblue', high = 'red4') +

  xlab('') + ylab('Enrichment score') +
  labs(
    title = 'GO Biological processes',
    subtitle = 'Top 30 terms ordered by Kolmogorov-Smirnov p-value',
    caption = 'Cut-off lines drawn at equivalents of p=0.05, p=0.01, p=0.001') +

  geom_hline(yintercept = c(-log10(0.05), -log10(0.01), -log10(0.001)),
    linetype = c("dotted", "longdash", "solid"),
    colour = c("black", "black", "black"),
    size = c(0.5, 1.5, 3)) +

  theme_bw(base_size = 24) +
  theme(
    legend.position = 'right',
    legend.background = element_rect(),
    plot.title = element_text(angle = 0, size = 16, face = 'bold', vjust = 1),
    plot.subtitle = element_text(angle = 0, size = 14, face = 'bold', vjust = 1),
    plot.caption = element_text(angle = 0, size = 12, face = 'bold', vjust = 1),

    axis.text.x = element_text(angle = 0, size = 12, face = 'bold', hjust = 1.10),
    axis.text.y = element_text(angle = 0, size = 12, face = 'bold', vjust = 0.5),
    axis.title = element_text(size = 12, face = 'bold'),
    axis.title.x = element_text(size = 12, face = 'bold'),
    axis.title.y = element_text(size = 12, face = 'bold'),
    axis.line = element_line(colour = 'black'),

    #Legend
    legend.key = element_blank(), # removes the border
    legend.key.size = unit(1, "cm"), # Sets overall area/size of the legend
    legend.text = element_text(size = 14, face = "bold"), # Text size
    title = element_text(size = 14, face = "bold")) +

  coord_flip()

showSigOfNodes(GOdata_MF, score(resultFisher), firstSigNodes = 5, useInfo = 'all')
```

## How significant GO terms are distributed over GO graph
```{r}
sel.go <- names(pValue.classic)[pValue.elim < pValue.classic]
 cbind(termStat(GOdata_MF, sel.go),
       elim = pValue.elim[sel.go],
       classic = pValue.classic[sel.go])
 pdf("plotGO.pdf")
 showSigOfNodes(GOdata_MF, score(resultElim), firstSigNodes = 5, useInfo = 'all')
 dev.off()
```

## Annotation 
nr. 5) in the vignete
```{r}
BPterms <- ls(GOBPTerm)
head(BPterms)

```

```{r working with TopGoData}
 a <- genes(GOdata_MF) ## obtain the list of genes
 head(a)
 numGenes(GOdata_MF)

 
 # retrieving a score of specified set of genes
selGenes <- sample(a, 10)
 gs <- geneScore(GOdata_MF, whichGenes = selGenes)
 print(gs)
 
 # unnamed vector or the scrore of all genes
 gs<-geneScore(GOdata_MF, whichGenes = selGenes, use.names=F)
 print(gs)
 gs <- geneScore(GOdata_MF, use.names = FALSE)
 str(gs)

#list of sig. genes
 sg <- sigGenes(GOdata_MF)
str(sg)
numSigGenes(GOdata_MF)

# updating the genes 
.geneList <- geneScore(GOdata_MF, use.names = TRUE)
 GOdata_MF ## more available genes
 GOdata_MF <- updateGenes(GOdata_MF, .geneList, topDiffGenes)
 GOdata_MF ## the available genes are now the feasible genes
 graph(GOdata_MF)                                              # ???is this number of nods7edges enough????
```


```{r}
ug <- usedGO(GOdata_MF)
 head(ug)
 
 sel.terms <- sample(usedGO(GOdata_MF), 10) #selecting 10 random GO terms to get the genes 
num.ann.genes <- countGenesInTerm(GOdata_MF, sel.terms) ## the number of annotated genes > num.ann.genes
 ann.genes <- genesInTerm(GOdata_MF, sel.terms) ## get the annotations
 head(ann.genes)
 
 #sel.term missing== you get all the GO terms
 ann.score <- scoresInTerm(GOdata_MF, sel.terms)
 head(ann.score)
 ann.score <- scoresInTerm(GOdata_MF, sel.terms, use.names = TRUE)
 head(ann.score)
 termStat(GOdata_MF, sel.terms)
```
´
## Analysing individual GO

```{r Analysing individual GO}
goID <- allRes[10, "GO.ID"]
gt <- printGenes(GOdata_MF, whichTerms = goID, chip = affyLib, numChar = 40)
### Error no value for BHLHE40? 
```




To do:

We cont with likelihood ratio test, ~ for the model use also other factors that influence- you need even number of comparison (6=6)
variance partition package https://rdrr.io/bioc/variancePartition/f/inst/doc/variancePartition.pdf 

doing heatmap using the genes (done)

age clust by similarity

- topGOtable
- pathways and functions - make a table 


#topgotable --> 4.0_pcaExplore_GO rmd.

```{r}
#using res, anno, PAS 
dds_GO <- DESeq(dds)
res<- results(dds_GO)
ENS_res<- merge(as.data.frame(res), KEY, by.y="gene_name", by.x= "row.names")
ENS_res$gene_id<-gsub("\\..*","",ENS_res$gene_id)
genenames_GO<-mapIds(org.Hs.eg.db,
                     keys=ENS_res$gene_id, 
                     column = "SYMBOL",
                     keytype = "ENSEMBL")
annotation_GO <- data.frame(gene_name = genenames_GO, #some duplication 
                                row.names = rownames(ENS_res),
                                stringsAsFactors = FALSE)
head(annotation_GO)  

anno_df_orgdb <- get_annotation_orgdb(dds = dds_GO,
                                      orgdb_species = "org.Hs.eg.db",
                                      idtype = "ENSEMBL")

anno_df_biomart <- get_annotation(dds = dds,
                                  biomart_dataset = "hsapiens_gene_ensembl",
                                  idtype = "ensembl_gene_id")


res$symbol<- mapIds(org.Hs.eg.db,
                            keys= row.names(res),
                            column="SYMBOL",
                            keytype="ENSEMBL",
                            multiVals="first")
res$entrez <- mapIds(org.Hs.eg.db, ##problem 
                            keys=row.names(res),
                            column="ENTREZID",
                            keytype="ENSEMBL",
                            multiVals="first")
resOrdered <- as.data.frame(res[order(res$padj),])
head(resOrdered)

de_df <- resOrdered[resOrdered$padj < .05 & !is.na(resOrdered$padj),]
de_symbols <- de_df$symbol
# extract background genes
bg_ids <- rownames(dds_GO)[rowSums(counts(dds_GO)) > 0]
bg_symbols <- mapIds(org.Hs.eg.db,
                     keys=bg_ids,
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")
# run the function
topgoDE_GO <- topGOtable(de_symbols, bg_symbols,
                             ontology = "BP",
                             mapping = "org.Hs.eg.db",
                             geneID = "symbol")

```

 