---
title: "MWE"
output: html_notebook
---
# Minimal Working Example GO-Terms

```{r}
library(tidyverse)
library(org.Hs.eg.db)
library(topGO)
library(pcaExplorer)
library(clusterProfiler)
library(ReactomePA)

# Required data: (From notebook: 1.0_metadata_exploration.rmd)
# (1) Test results PAS vs. Ctrl and 
# (2) CtrlMus vs. Ctrl, both unfiltered!

results_PAS_ctrl
results_mus_ctrl
```

## (1) Prepare input data
First subset for genes with padj < 0.05 in both test results. Remove significant genes in CtrlMus vs. Ctrl from PAS vs. Ctrl.
```{r}
# Get significant genes to remove from CtrlMus vs. Ctrl comparison
rm <- results_mus_ctrl |>
  subset(padj < 0.05) |>
  rownames_to_column(var = "gene_id") |>
  pull(gene_id)

dat <- results_PAS_ctrl |>
  rownames_to_column(var = "gene_id")|>
  subset(!gene_id %in% rm)

# Define significance threshold and divide into up and down genes based on logFC
genes <- dat |> subset(padj < 0.05) |> pull(gene_id)
up_genes <- dat |> subset(padj < 0.05 & dat$log2FoldChange > 0) |> pull(gene_id)
dn_genes <- dat |> subset(padj < 0.05 & dat$log2FoldChange < 0) |> pull(gene_id)
all_genes <- dat |> pull(gene_id)

```

## (2) Run TopGO with Biological (BP) process and Molecular function (MF)
```{r}
upRes <- lapply(c("BP", "MF"), function(ontology) {
  topgoDE <- topGOtable(DEgenes = up_genes,
                        BGgenes = all_genes,
                        ontology = ontology,
                        mapping = "org.Hs.eg.db",
                        geneID = "ENSEMBL")
  })
upRes
```


## (3) Run other ORA tests
```{r}
# Add ENTREZ IDs
ensembl_to_entrez <- AnnotationDbi::select(org.Hs.eg.db, unique(dat$gene_id),
                                           keytype = "ENSEMBL",
                                           columns = c("ENTREZID"))

dat <- dat |> inner_join(ensembl_to_entrez, by = c("gene_id" = "ENSEMBL"))

universe <- unique(dat$ENTREZID)
genes <- dat |> subset(padj < 0.05) |> pull(ENTREZID) |> unique()
```

```{r}
# KEGG pathway over-representation analysis
Kegg <- enrichKEGG(
  gene = genes,
  universe = universe,
  organism = 'hsa',
  pvalueCutoff = 0.05)
```
```{r}
Reactome <- enrichPathway(
  gene = genes,
  organism = "human",
  universe = universe,
  pvalueCutoff = 0.05,
  readable = TRUE)

as_tibble(Reactome@result)
```

```{r}
WikiPathway <- enrichWP(
  gene = genes,
  universe = universe,
  organism = 'Homo sapiens',
  pvalueCutoff = 0.05)

as_tibble(WikiPathway@result)
```

## Finer control on GO term tests

```{r}
ontology <- "BP"
#ontology <- "CC" #(Cellular Component)
#ontology <- "MF" #(Molecular Function)

# Decide test algorithm
algorithm <- "weight01"
#algorithm <- "classic"
#algorithm <- "elim"
#algorithm <- "weight"
#algorithm <- "lea"
#algorithm <- "parentchild"

# Define the statistical test used
# For tests based on gene counts
statistic <- "fisher"      # Fischer's exact test
#For tests based on gene scores or gene ranks
#statistic <- "ks"         # Kolmogorov-Smirnov test
#statistic <- "t"          # t-test
# For tests based on gene expression
#statistic <- "globaltest" # globaltest


# Produce list of up's
upList <- factor(as.integer(all_genes %in% up_genes))
names(upList) <- all_genes

upGOdata <- new("topGOdata", ontology = ontology, allGenes = upList, geneSel = function(x)(x == 1), 
                nodeSize = 10, annot = annFUN.org, mapping = "org.Hs.eg.db", ID = "ENSEMBL")

upRes <- runTest(upGOdata, algorithm = algorithm, statistic = statistic)

# Create result table
# Build up result table
up_tab <- GenTable(upGOdata, Pval = upRes, topNodes = 20)

# Update table with full GO term name
# Update up table 
up_tab$Term <- sapply(up_tab$"GO.ID", function(go) Term(GO.db::GOTERM[[go]]))

up_sigGenes <- sigGenes(upGOdata)

# Retrieve gene symbols for each GO from the test result
up_AnnoList <- lapply(up_tab$"GO.ID", 
                      function(x) as.character(unlist(genesInTerm(object = upGOdata, whichGO = x))))

up_SigList <- lapply(up_AnnoList, function(x) intersect(x, up_sigGenes))

#Coerce gene list to a comma-separated vector
up_tab$Genes <- sapply(up_SigList, paste, collapse = ",")

# Cbind first 5 up 
cbind(head(up_tab$Genes, 5))
```